Design and Implementation of\\Anglican Probabilistic Programming Language

### 2016-09-01 slides/slides.tex
in progress

\begin{frame}
\titlepage
\vfill
\center
Paper and slides: \url{https://bitbucket.org/probprog/anglican-white-paper}

@@

\section{Why probabilistic programming?}

@@

\begin{frame}{Definition}

A \textbf{probabilistic program} is a stateful deterministic
computation $\mathcal{P}$:
\pause
\begin{itemize}
\item Initially, $\mathcal{P}$ expects no arguments.
    \pause
\item On every call, $\mathcal{P}$ returns
    \begin{itemize}
        \item a distribution $F$,
        \item a distribution and a value $(G, y)$,
        \item a value $z$, 
        \item or $\bot$.
    \end{itemize}
    \pause
\item Upon returning $F$, $\mathcal{P}$ expects $x\sim F$.
    \pause
\item Upon returning $\bot$, $\mathcal{P}$ terminates.
\end{itemize}
\pause
A program is run by calling $\mathcal{P}$ repeatedly until termination.

The probability of each \textbf{trace} is $\propto\prod_{i=1}^{\left|\pmb{x}\right|} p_{F_i}(x_i) \prod_{j=1}^{\left|\pmb{y}\right|}p_{G_j}(y_{j})$.
\end{frame}

\begin{frame}{Evolution of Compiler Development}
    \begin{itemize} 
        \item 1950s: Ask a programmer to implement an algorithm efficiently: They'll write it on their own in assembly.
            \begin{itemize}
                \item   No good compilers; problem-dependent optimizations that only human expert could see.
            \end{itemize}
            \pause
        \item   1970s: Novice programmers use high-level languages and let compiler work out details, experts still write assembly.
            \begin{itemize}
                \item   Experts still write custom assembly when speed critical.
            \end{itemize}
            \pause
        \item   2000s: On most problems, even experts can't write faster assembly than optimizing compilers.
            \begin{itemize}
                \item   can automatically profile (JIT).
                \item   can take advantage of paralellization, complicated hardware, make appropriate choices w.r.t. caching.
                \item   Compilers embody decades of compiler research
            \end{itemize}
    \end{itemize}
\end{frame}

\begin{frame}{Evolution of Inference}

    \begin{itemize}
\item           2010: Ask a grad student to implement inference efficiently: They'll write it on their own.
    \begin{itemize}
\item   No good automatic inference engines; problem-dependent optimizations that only human expert can see.
    \end{itemize}
    \pause
\item   2015: Novice grad students use automatic inference engines and let compiler work out details, experts still write their own inference.
    \begin{itemize}
\item   Experts still write custom inference when speed critical.
    \end{itemize}
    \pause
\item   2020: On most problems, even experts can't write faster inference than mature automatic inference engines.
    \begin{itemize}
\item   Can use paralellization, sophisticated hardware
\item   Can automatically choose appropriate methods (meta-reasoning?).
\item   Inference engines will embody 1 decade (!) of PP research.
    \end{itemize}
    \end{itemize}
\end{frame}

\section{Why functional?}

@@

\begin{frame}{Importance Sampling}
    \begin{algorithmic}
        \Loop
            \State Run $\mathcal{P}$, computing weight $w=\prod_{j=1}^{\left|\pmb{y}\right|}p_{G_j}(y_{j})$.
            \State output $\pmb{z}, w$.
        \EndLoop
    \end{algorithmic}
    \begin{itemize}
        \item[] 
         \item Simple --- good.
         \item Slow convergence (unless one knows the answer) --- bad.
     \end{itemize}
     \vfill
     Can we do better?
\end{frame}

\begin{frame}{Lightweight Metropolis-Hastings (LMH)}
\begin{algorithmic}
    \State {Run $\mathcal{P}$ once, remember $\pmb{x}, \pmb{z}$.}
    \Loop
         \State {Uniformly select $x_i$}.
         \State {Propose a value for $x_i$.}
         \State {Run $\mathcal{P}$, remember $\pmb{x'}, \pmb{z'}$}.
         \State {Accept ($\pmb{x,z}=\pmb{x',z'}$) or reject with MH probability}.
         \State {Output $\pmb{z}$.}
    \EndLoop
\end{algorithmic}
\vfill
Can we do better?
\begin{itemize}
    \item Particle Markov Chain Monte Carlo
    \item Variational Inference
    \item ...

\end{itemize}
\end{frame}

\begin{frame}{Challenges}
    \begin{itemize}
        \item `Transformational compilation' --- limited languages, some less ugly than others.
        \item Slow inference --- Markov Chain Monte Carlo is \textbf{always slow} (but there are Hybrid Monte Carlo, Variational Inference, ...)
        \item How to learn persistently --- what is the learned model? Transfer learning?
        \item Handling indeterminism (learning policies).
    \end{itemize}
\end{frame}

>>>>

\begin{frame}
\titlepage
\vfill
\center
\url{https://bitbucket.org/probprog/anglican-white-paper}\\
\url{https://bitbucket.org/probprog/anglican} \\
\url{http://www.robots.ox.ac.uk/~fwood/anglican/index.html}

@@

\section{Motivation}

@@

\begin{frame}{More examples}
    \begin{itemize}
        \item \textit{Intruder detection} --- 
            given a log of \textbf{times} and \textbf{amounts} of
            payments in a bank account, how likely that the baccount was
            compromised?
        \pause
        \item \textit{Counterfactual reasoning} --- There are
            \textbf{two routes} from Jerusalem to Tel Aviv:
            1 and 443.  Based on traffic reports, I chose route
            1 and was late. Would I arrive on time If I chose 443
            instead?
        \pause 
        \item \textit{(Due to Stuart Russell)} If you observe that a
            student GPA is exactly 4.0 in a model of
            transcripts of students from the USA (GPA's from
            0.0 to 4.0) and India (GPA's from 0.0 to
            10.0) what is the probability that the student
            is from India? 
    \end{itemize}
\end{frame}

@@

\begin{frame}{Importance Sampling}
    \begin{algorithmic}
        \Loop
            \State Run program, computing weight based on observations.
            \State Output result and weight.
        \EndLoop
    \end{algorithmic}
    \begin{itemize}
        \item[] 
         \item Simple --- good.
         \item Slow convergence (unless one knows the answer) --- bad.
     \end{itemize}
     \vfill
     Can we do better?
\end{frame}

\begin{frame}{Lightweight Metropolis-Hastings (LMH)}
\begin{algorithmic}
    \State Run program once, remembering random choices.
    \Loop
         \State Uniformly select one random choice.
         \State Propose a new value for the choice.
         \State Re-run the program.
         \State Accept or reject with MH probability.
         \State Output result.
    \EndLoop
\end{algorithmic}
\vfill
Can we do better?
\begin{itemize}
    \item Particle methods
    \item Variational inference
    \item ...
\end{itemize}
\end{frame}

\begin{frame}{Why functional?}
    We want a functional language because an inference algorithm
    controls the execution:
    \begin{itemize}
        \item A program is run many (often many hundreds of
            thousands) of times (with almost any algorithm).
        \item A program must be partially re-executed multiple
            times from different positions (particle methods).
        \item We want to reason about the distribution defined
            by the program.
    \end{itemize}
\end{frame}

\begin{frame}{Why Closure?}
\begin{itemize}
    \item Runs on JVM --- easy deployment and access to libraries.
        \pause
    \item A Lisp --- we (ab)use the macro facility.
        \pause
    \item Church (\url{https://en.wikipedia.org/wiki/Church_(programming_language)}) is derived from Scheme.
\end{itemize}
\pause
Others use:
    \begin{itemize}
        \item Scheme --- (Church, Venture).
        \item Scala --- Figaro.
        \item Haskell --- Hakaru, Model-Bayes.
        \item ...
    \end{itemize}
    \pause
    As well as Python, C\#,  and other languages.
\end{frame}

\section{Design Outline}

\subsection{Language}

\subsection{Macro-based compilation}

\subsection{Implementation Highlights}

\subsection{Managing stack size}

\subsection{Probabilistic forms}

\subsection{Memoization}

\section{Inference Algorithms}

\section{Definitions and Runtime Library}
